
---
title: "Reproducible Research: Peer Assessment 1"
author: "Marta Kaczmarz"
date: "Sunday, September 14, 2014"
output: 
  html_document:
    keep_md: true
---

This is a report (R mark document) that answers the questions from Peer Assessment 1. 


## Loading and preprocessing the data

To load the dataset stored in a comma-separated-value (CSV) file you can use ```read.csv()``` function:
```{r}
act.data <- read.csv('activity.csv', sep = ",")
```
Dataset has 17,568 observations and 3 variables. I also use function ```head()``` to see the first parts of a data frame or function:
```{r}
head(act.data)
```
In my opinion everything looks correct so data do not need transformation.

## What is mean total number of steps taken per day?

To answer to the question: *What is mean total number of steps taken per day?* first I prepared the data to make a histogram of the total number of steps taken each day:
```{r}
tot.no.steps <- tapply(act.data$steps,act.data$date, sum , na.rm = TRUE)
```
and then plot a histogram:
```{r}
hist(tot.no.steps, main = "Hisogram of the total number of steps", 
      xlab = "number of steps taken each day", col = "blue")
```

Then I calculated the mean and median total number of steps taken per day:
```{r}
mean(tot.no.steps, na.rm = "TRUE")
median(tot.no.steps, na.rm = "TRUE")
```

## What is the average daily activity pattern?

I made a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days with R package ```ggplot2```:
```{r}
avg.steps <- tapply(act.data$steps,act.data$interval, mean , na.rm = TRUE)
avg.steps <- data.frame(interval = as.numeric(names(avg.steps)),mean.steps = as.vector(avg.steps))

library(ggplot2)
ggplot(avg.steps,aes(interval ,mean.steps)) + geom_line()+
  xlab('5-minute interval') +
  ylab('avg. number of steps')
```
  
Next
I tried to answer to the qestion *Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*:
```{r}
avg.steps[which.max(avg.steps$mean.steps),]
```
So the answer is 835.

## Imputing missing values

The total number of missing values in the dataset:
```{r}
sapply(act.data, function(x) sum(is.na(x)))
```

The other options are: ```apply()``` or ```colSums()``` functions:
```{r}
apply(is.na(act.data),2,sum)
colSums(is.na(act.data))
```

I used a ceiling of the mean for that 5-minute interval to filling in all of the missing values in the dataset:
```{r}
merged.data <- merge(act.data, avg.steps[,],  by.x = "interval", by.y = "interval", sort = FALSE)
new.data <- merged.data[order(merged.data[,3], merged.data[,2],merged.data[,1]), ]
```

The next step was to create a new dataset that is equal to the original dataset but with the missing data filled in:
```{r}
new.data$steps[is.na(new.data[,2])] <- ceiling(new.data$mean.steps[is.na(new.data[,2])])

head(new.data)
```

I made again a histogram of the total number of steps taken each day and cCalculated the mean and median total number of steps taken per day for filling data:

```{r}
tot.no.steps.new <- tapply(new.data$steps, new.data$date, sum , na.rm = TRUE)

hist(tot.no.steps.new, main = "Hisogram of the total number of steps", 
      xlab = "number of steps taken each day", col = "blue")
```

Mean and median total number of steps taken per day:
```{r}
mean(tot.no.steps.new, na.rm = "TRUE")
median(tot.no.steps.new, na.rm = "TRUE")
```

These values are differ from the estimates from the first part of the assignment, to make better compare:
```{r}
data.frame(Na.filling = c("NO", "YES"),
            mean = c(mean(tot.no.steps, na.rm = "TRUE"),mean(tot.no.steps.new, na.rm = "TRUE")),
            median = c(median(tot.no.steps, na.rm = "TRUE"),median(tot.no.steps.new, na.rm = "TRUE")) 
           )
```
Imputing missing data on the estimates of the total daily number of steps cause increasing of mean and median this value.


## Are there differences in activity patterns between weekdays and weekends?

I created a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day:
```{r}
new.data$weekdays <- weekdays(as.Date(new.data$date))
new.data$weekdays[new.data$weekdays %in% c("niedziela", "sobota")]<- "weekend"
new.data$weekdays[!new.data$weekdays %in% c("weekend")] <- "weekday"

head(new.data)
table(new.data$weekdays)
```

To make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days, first I prepered a data separately for "weekday" and "weekend":

```{r}
avg.steps.weekday <- tapply(new.data$steps[new.data$weekdays == "weekday"],new.data$interval[new.data$weekdays == "weekday"], mean)
avg.steps.weekday <- data.frame(interval = as.numeric(names(avg.steps.weekday)),mean.steps = as.vector(avg.steps.weekday))

avg.steps.weekend <- tapply(new.data$steps[new.data$weekdays != "weekday"],new.data$interval[new.data$weekdays != "weekday"], mean)
avg.steps.weekend <- data.frame(interval = as.numeric(names(avg.steps.weekend)),mean.steps = as.vector(avg.steps.weekend))
```

Finaly plot:

```{r, fig.heigh = 5}
par(mfrow = c(2,1))
plot(avg.steps.weekday, type = "l", main = "weekday", ylab = "Number of steps")
plot(avg.steps.weekend, type = "l", main = "weekend",ylab = "Number of steps")
```
